---
title: "Tracking Loggerhead Turtle Carapace Length"
output: html_notebook
---

```{r}
library(tidyverse)
library(ggprism)
library(ggplot2)
```

```{r}
loggerhead_data <- read_csv("Loggerhead_oceanic_stage_duration.csv")

loggerhead_df <- loggerhead_data %>%
  mutate(Sex = ifelse(Sex == "unknown", NA, Sex)) %>% 
  drop_na()

loggerhead_df

```


```{r}
loggerhead_df %>%
    ggplot() + 
  geom_boxplot(mapping = aes(x = Sex , y = Straight_Carapace_Length, fill = Sex)) +
  theme_prism() +
  labs(
    title = "Sex Influences Carapace Length", 
    subtitle = "On average, males have longer carapace length than females",
    x = "Sex",
    y = "Carapace Length (in)"
  ) +
  stat_summary(mapping = aes(x = Sex , y = Straight_Carapace_Length), 
    fun = "mean"
  )

```


### check normality of the variables to determine if the data is parametric or non-parametric

```{r}
loggerhead_df %>%
  filter(Sex == "female") %>% 
  ggplot() +
  geom_qq( mapping = aes(sample = Straight_Carapace_Length)) +
  geom_qq_line(mapping = aes(sample = Straight_Carapace_Length)) + 
  theme_prism() + 
  labs(
    title = "Normality of Female Carapace Length Distributions"
  )

loggerhead_df %>%
  filter(Sex == "male") %>% 
  ggplot() +
  geom_qq( mapping = aes(sample = Straight_Carapace_Length)) +
  geom_qq_line(mapping = aes(sample = Straight_Carapace_Length)) +
  theme_prism() + 
  labs(
    title = "Normality of Male Carapace Length Distributions"
  )
  
```

## does not look very normal so lets do a nonparametric test 

```{r}
F_length <- loggerhead_df  %>% filter(Sex == "female") %>% pull(Straight_Carapace_Length)
M_length <- loggerhead_df %>% filter(Sex == "male") %>% pull(Straight_Carapace_Length)
wilcox.test(F_length, M_length)
```
### we reject the null (p is less than 0.05)

## INFERENTIAL STATS

## Resampling

#### For every resample, we take the mean carapace length of females and subtract it from the mean carapace length of males.

```{r}
hundred_resamples <- c()

for(i in 1:100){
  oneresample <- loggerhead_df %>%
    sample_n(size=333, replace = TRUE)
  
 sc_length <- oneresample %>% 
  group_by(Sex) %>%
  summarize(mean_length = mean(Straight_Carapace_Length))
  
test_stat <- sc_length %>% 
  filter(Sex == "male") %>% 
  pull(mean_length) -
  sc_length %>%
  filter(Sex == "female") %>%
  pull(mean_length)

hundred_resamples <- hundred_resamples %>% rbind(test_stat)
}

```

## convert the vector into a data frame 
```{r}
hundred_resamples <- enframe(hundred_resamples)
```


## Find standard error 
```{r}
SE <- hundred_resamples %>% pull(value) %>% sd()
SE
```

## This point estimate is the average of lengths between sexes, estimating the true difference in length of the population
```{r}
summary_turtles <- loggerhead_df %>% 
  group_by(Sex) %>%
  summarise(mean_length = mean(Straight_Carapace_Length))
PE <- summary_turtles %>% 
  filter(Sex == "male") %>% pull(mean_length) -
  summary_turtles %>%
  filter(Sex == "female") %>% pull(mean_length)

PE
```

## The confidence interval is the range of plausible differences in length between male and female loggerheads 
```{r}
lower_bound_CI <- PE - (1.96 * SE) 
upper_bound_CI <- PE + (1.96 * SE)

lower_bound_CI
upper_bound_CI
```


## graphing bootstrapping, CI, PE

```{r}
ggplot(hundred_resamples, aes(x = value)) +
  geom_histogram(binwidth = 0.1, fill = "lightblue", color = "white") +
  # shaded CI region
  annotate("rect",
           xmin = lower_bound_CI, xmax = upper_bound_CI,
           ymin = 0,            ymax = Inf,
           alpha = 0.2,         fill = "blue") +
  # dashed lines at the CI bounds
  geom_vline(xintercept = lower_bound_CI, linetype = "dashed", color = "blue", size = 1) +
  geom_vline(xintercept = upper_bound_CI, linetype = "dashed", color = "blue", size = 1) +
  # solid line at the point estimate
  geom_vline(xintercept = PE, color = "red", size = 1.2) +
  theme_prism() +
  labs(
    title =  "Bootstrap Distribution of Mean Carapace Length Difference",
    subtitle = sprintf("95%% CI: [%.2f, %.2f]    Point Estimate: %.2f", 
                       lower_bound_CI, upper_bound_CI, PE),
    x = "Difference in Mean Carapace Length (in) (male – female)",
    y = "Frequency"
  )
```
## from these data, we estimate males are, on average 6.3 (the point estimate) inches larger than females. If we were to repeat this experiment, we'd expect males to be between 2.49 and 10.11 inches (the confidence interval) larger than females. 